# Description {.unnumbered}

## Objective

The objective of this repository is to flexibly simulate de novo datasets with different missingness scenarios. The simulated datasets approximate the distributions of the empirical cohort used in the [`HDMI` project](https://gitlab-scm.partners.org/drugepi/hd-mi).

::: {.content-visible when-format="html"}
## Dependencies {#sec-dependencies .unnumbered}

This is a quarto project and R package dependencies are managed through the `renv` package. All packages and their versions can be viewed in the lockfile `renv.lock`. All required packages and the appropriate versions can be installed by running the following command:

```{r, eval=FALSE, echo=TRUE}
renv::restore()
```

::: callout-important
The dependencies are managed through Posit's repository package manager (RSPM) version for your underlying OS distribution. If you use a different operating system than MacOS, please head over to the RSPM setup website and follow these steps:

1.  Go to the [RSPM setup website](https://packagemanager.posit.co/client/#/repos/cran/setup?distribution=redhat-9)

2.  Choose the operating system (if Linux, also choose the Linux distribution)

3.  Go to **`Repository URL:`** and copy-paste the URL to the options statement in the `.Rprofile` file

    *options(repos = c(REPO_NAME = "URL"))*
:::

## Reproducibility {.unnumbered}

Follow these steps in RStudio to reproduce this study:

::: callout-note
1.  Install all necessary dependencies (see above)
2.  In RStudio, run all scripts via `quarto render` or `Build > Render Book` (make sure quarto is installed)
:::

## Repository structure and files {.unnumbered}

### Directory overview {.unnumbered}

```{r, echo=FALSE}
fs::dir_tree(recurse = 1)
```

-   .Rprofile - defines paths, activates `renv`, options for Posit R package manager
-   scripts - main R/RMarkdown scripts
-   functions - helper functions called in scripts
-   renv/renv.lock - `renv` directories to manage R dependencies and versions used in this simulation
-   docs - output of Quarto scripts
-   README - essential information about the project (README.Rmd renders to README.md via update_README.R after each `quarto render` command)

:::



